Fundamental Problems with the Evidence Base of Adolescent Depression Treatments

### 2024-03-20 Untitled.qmd
Updating untitled.qmd doc to fix some formatting problems. Playing around with rendering to see if I can get it to work in PDF.

---
title: "Fundamental Problems with the Evidence Base of Adolescent Depression Treatments"
author: "Stringaris et al"
prefer-html: true
format: docx
editor: visual
---

@@

---

---

@@

When taking only those studies where the active arm was either CBT, fluoxetine or escitalopram, there was a significant difference between psychotherapy and medication studies. This difference did not reach statistical significance when excluding psychotherapy studies that recruited samples with sub-clinical depression or when excluding studies that used wait list as their control. Please refer to Table XXXXX for these results. <!-- Charlotte, please point to the Supplementary Table where these results live -->

@@

As can be seen in Table 1, psychotherapy trials featured a significantly higher percentage of women when compared to medication trials. On average, samples were comprised of `r round(results_met_perc_women_overall$TE_random[results_met_perc_women_overall$subgroup == 1][[2]],1)`% (*SE* = `r round(results_met_perc_women_overall$seTE_random[results_met_perc_women_overall$subgroup == 1][[2]],1)` women across psychotherapy trials and `r round(results_met_perc_women_overall$TE_random[results_met_perc_women_overall$subgroup == 0][[2]],1)`% (*SE* = `r round(results_met_perc_women_overall$seTE_random[results_met_perc_women_overall$subgroup == 0][[2]],1)` women across medication trials. Excluding sub-clinical and waitlist control studies yielded very similar results. This difference in gender composition however did not reach statistical significance when only CBT, fluoxetine and escitalopram trials were included . <!-- Charlotte please refer to Supple Table -->

@@

As can be seen in Table 1, mean age was `r round(results_met_age$TE_random[results_met_age$subgroup == 1][[2]],2)` (*SE* = `r round(results_met_age$seTE_random[results_met_age$subgroup == 1][[2]],2)`) across psychotherapy trials and `r round(results_met_age$TE_random[results_met_age$subgroup == 0][[2]],2)` (*SE* = `r round(results_met_age$seTE_random[results_met_age$subgroup == 0][[2]],2)`) across medication trials, with no significant between group differences. There were no significant differences in mean age between modalities on further sensitivity analyses.

@@

```{r, warning=FALSE, message = FALSE, echo = FALSE}
#df_overall_baseline_results

df_overall_baseline_results %>%
  kbl(align = "l", caption = "Table 1: Sample characteristics at baseline across medication and psychotherapy studies", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>%
  row_spec(row = 1, bold = T) %>%
  row_spec(c(4, 7), bold = T, extra_css = "border-top: 1px solid black;") %>%
  column_spec(1, border_right = T) %>%
  add_indent(c(2, 3, 5, 6, 8, 9)) %>%
  footnote(number = c("Statistic derived from the test for subgroup differences in the relevant meta-analysis"))
```

@@

ggplot(df_baseline_demographics, aes(x = factor(psy_or_med), y = sites, color = factor(psy_or_med))) +
  geom_point(position = position_jitter(width = 0.2), size = 2) +
  labs(title = "Number of Sites for Psychotherapy and Medication Trials",
       x = "Treatment Modality",
       y = "Number of Sites") +
  scale_color_discrete(name = "Treatment Type", labels = c("Medication", "Psychotherapy"))

### Number of trial sites

There was a significant difference between the number of sites in medication versus psychotherapy trials, as can be seen in Tables 2 and 3. Average number of trial sites was significantly higher across medication studies ( *M* = `r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Medication"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Medication"]`compared to psychotherapy studies ( *M* =`r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Psychotherapy"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Psychotherapy"]`(*t* (`r sites_t_test$parameter`) = `r round(sites_t_test$statistic,2)`, *p* =`r round(sites_t_test$p.value,3)`).

df_sitest_summary_stats <- df_sites_summary_stats %>%
  kbl(caption = " Number of sites across medication and psychotherapy studies") %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB") %>%
  column_spec(column = 1, border_right = T)

@@

As can be seen in Figure 1, there were substantial differences between the four arms of the metanalysis with striking differences between the medication and the psychotherapy control arms. In particular, placebo had an SMD = `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "medication_control", ]$coef_means, 2)`, 95% CI: `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "medication_control", ]$lower_ci, 2)` to `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "medication_control", ]$upper_ci, 2)`, whereas psychotherapy controls had an SMD = `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "psychotherapy_control", ]$coef_means, 2)` , 95% CI: `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "psychotherapy_control", ]$lower_ci, 2)` to `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "psychotherapy_control", ]$upper_ci, 2)` (see Table 4 below).

@@

![Figure 1.](plot_means_all.pdf){alt="Figure 1."}

```{r, warning=FALSE, message = FALSE, echo=FALSE}
# df_mean_coefs_from_sim
df_mean_coefs_from_sim %>%
  kbl(align = "l", caption = " Summary statistics of the estimated SMDs") %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB") %>%
  column_spec(column = 1, border_right = T)

@@

```{r, warning=FALSE, message = FALSE, echo = FALSE}
#aggregate_results_overall
aggregate_results_overall %>%
  kbl(caption = "Table 5: Meta-regression results") %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB") %>%
  column_spec(column = 1, border_right = T)

@@

```{r, warning=FALSE, message = FALSE, echo=FALSE}
# We need to read in the appropriate dataset which describes the active and control conditions used in
# psychotherapy trials.

@@

colnames(df_descr_psy_conditions)

@@

control_summary %>%
  kbl(align = "l", caption = "Table X: Comparing the intensity of the intervention between active and control arms of psychotherapy studies", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>% 
  column_spec(1, border_right = T) %>% 
  pack_rows("Number of sessions", 1, 2) %>% 
  pack_rows("Frequency (weeks)", 3, 4) %>% 
  pack_rows("Session length (mins)", 5, 6) %>%
  pack_rows("Total intervention hours", 7, 8) %>%
  pack_rows("Total intervention period (weeks)", 9, 10)

@@

colnames(df_overall_baseline_results)[8] <- paste0(colnames(df_overall_baseline_results)[8], footnote_marker_number(1))
```

```{r, warning=FALSE, message = FALSE}
control_summary_t_tests %>%
  kbl(align = "l", caption = "Table X: Results for t-tests comparing intervention intensity between active and control arms of psychotherapy trials") %>% 
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>% 
  row_spec(row = 0, background = "#EBEBEB") %>% 
  column_spec(column = 1, border_right = T)

@@

Overall, the active arms of psychotherapy studies were considerably more intensive than the control arms they were compared against (see Table X for summary statistics). Active intervention arms featured significantly more sessions than control arms (*p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Number of sessions"]`). These sessions were also significantly longer (*p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Session length (mins)"]`) and more frequent (*p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Frequency (weeks)"]`). The total hours involved in an intervention were higher in active vs control arms (*p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Total intervention hours"]`) though the total period of the intervention was similar (*p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Total intervention period (weeks)"]`). <!-- Charlotte, we need to have all the t-values and dfs for the t-tests -->

@@

```{r, warning=FALSE, message = FALSE}
df_severity_sens %>%
  kbl(align = "l", caption = "Table X: Sensitivity analyses for comparison of baseline severity levels between psychotherapy and medication trials", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>% 
  row_spec(row = 1, bold = T) %>% 
  row_spec(c(4, 7), bold = T, extra_css = "border-top: 1px solid black;") %>%
  column_spec(1, border_right = T) %>% 
  add_indent(c(2, 3, 5, 6, 8, 9)) %>% 
  footnote(number = c("Statistic derived from the test for subgroup differences in the relevant meta-analysis"))
```

@@

```{r, warning=FALSE, message = FALSE}
df_gender_sens %>%
  kbl(align = "l", caption = "Table X: Sensitivity analyses for comparison of gender composition between psychotherapy and medication trials", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>% 
  row_spec(row = 1, bold = T) %>% 
  row_spec(c(4, 7), bold = T, extra_css = "border-top: 1px solid black;") %>%
  column_spec(1, border_right = T) %>% 
  add_indent(c(2, 3, 5, 6, 8, 9)) %>% 
  footnote(number = c("Statistic derived from the test for subgroup differences in the relevant meta-analysis"))
```

@@

```{r, warning=FALSE, message = FALSE}
df_age_sens %>%
  kbl(align = "l", caption = "Table X: Sensitivity analyses for comparison of age between psychotherapy and medication trials", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>% 
  row_spec(row = 1, bold = T) %>% 
  row_spec(c(4, 7), bold = T, extra_css = "border-top: 1px solid black;") %>%
  column_spec(1, border_right = T) %>% 
  add_indent(c(2, 3, 5, 6, 8, 9)) %>% 
  footnote(number = c("Statistic derived from the test for subgroup differences in the relevant meta-analysis"))
```

>>>>

---
title: "Fundamental Problems with the Evidence Base of Adolescent Depression Treatments"
author: "Stringaris et al"
output:
  quarto::pdf_document:
    pandoc_args: --pdf-engine=pdflatex
---

@@

------------------------------------------------------------------------

------------------------------------------------------------------------

@@

When taking only those studies where the active arm was either CBT, fluoxetine or escitalopram, there was a significant difference between psychotherapy and medication studies. This difference did not reach statistical significance when excluding psychotherapy studies that recruited samples with sub-clinical depression or when excluding studies that used wait list as their control. Please refer to Table S1 for these results.

@@

As can be seen in Table 1, psychotherapy trials featured a significantly higher percentage of women when compared to medication trials. On average, samples were comprised of `r round(results_met_perc_women_overall$TE_random[results_met_perc_women_overall$subgroup == 1][[2]],1)`% (*SE* = `r round(results_met_perc_women_overall$seTE_random[results_met_perc_women_overall$subgroup == 1][[2]],1)`) women across psychotherapy trials and `r round(results_met_perc_women_overall$TE_random[results_met_perc_women_overall$subgroup == 0][[2]],1)`% (*SE* = `r round(results_met_perc_women_overall$seTE_random[results_met_perc_women_overall$subgroup == 0][[2]],1)`) women across medication trials. Excluding sub-clinical and waitlist control studies yielded very similar results. This difference in gender composition however did not reach statistical significance when only CBT, fluoxetine and escitalopram trials were included (see Table S2).

@@

As can be seen in Table 1, mean age was `r round(results_met_age$TE_random[results_met_age$subgroup == 1][[2]],2)` (*SE* = `r round(results_met_age$seTE_random[results_met_age$subgroup == 1][[2]],2)`) across psychotherapy trials and `r round(results_met_age$TE_random[results_met_age$subgroup == 0][[2]],2)` (*SE* = `r round(results_met_age$seTE_random[results_met_age$subgroup == 0][[2]],2)`) across medication trials, with no significant between group differences. There were no significant differences in mean age between modalities on further sensitivity analyses (see Table S3).

@@

df_overall_baseline_results %>%
  mutate_all(~if_else(is.na(.), "", as.character(.))) %>% 
  kbl(align = "l", caption = "Sample characteristics at baseline across medication and psychotherapy studies", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>%
  row_spec(row = 1, bold = T) %>%
  row_spec(c(4, 7), bold = T, extra_css = "border-top: 1px solid black;") %>%
  column_spec(1, border_right = T) %>%
  add_indent(c(2, 3, 5, 6, 8, 9)) %>%
  footnote(number = c("Statistic derived from the test for subgroup differences in the relevant meta-analysis"))

```{r, warning=FALSE, message = FALSE, echo=FALSE, results='hide'}

@@

# ggplot(df_baseline_demographics, aes(x = factor(psy_or_med), y = sites, color = factor(psy_or_med))) +
#   geom_point(position = position_jitter(width = 0.2), size = 2) +
#   labs(title = "Number of Sites for Psychotherapy and Medication Trials",
#        x = "Treatment Modality",
#        y = "Number of Sites") +
#   scale_color_discrete(name = "Treatment Type", labels = c("Medication", "Psychotherapy"))

```
### Number of trial sites

There was a significant difference between the number of sites in medication versus psychotherapy trials, as can be seen in Tables 2 and 3. Average number of trial sites was significantly higher across medication studies ( *M* = `r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Medication"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Medication"]`) compared to psychotherapy studies ( *M* =`r df_sites_summary_stats$Mean[df_sites_summary_stats$Modality == "Psychotherapy"]`, *SD* =`r df_sites_summary_stats$SD[df_sites_summary_stats$Modality == "Psychotherapy"]`)(*t* (`r round (sites_t_test$parameter, 2)`) = `r round(sites_t_test$statistic,2)`, *p* =`r sites_t_test$p.value`).

```{r, warning=FALSE, message = FALSE, echo = FALSE, asis = TRUE }

df_sites_summary_stats <- df_sites_summary_stats %>%
  kbl(caption = "Number of sites across medication and psychotherapy studies") %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB") %>%
  column_spec(column = 1, border_right = T)

@@

As can be seen in Figure 1, there were substantial differences between the four arms of the metanalysis with striking differences between the medication and the psychotherapy control arms. In particular, placebo had an SMD = `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "medication_control", ]$coef_means, 2)` (95% CI: `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "medication_control", ]$lower_ci, 2)` to `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "medication_control", ]$upper_ci, 2)`) whereas psychotherapy controls had an SMD = `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "psychotherapy_control", ]$coef_means, 2)` (95% CI: `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "psychotherapy_control", ]$lower_ci, 2)` to `r round(df_mean_coefs_from_sim[df_mean_coefs_from_sim$condition == "psychotherapy_control", ]$upper_ci, 2)` ) (see Table 4 below).

@@

![Figure 1.](plot_means_all.pdf)

```{r, warning=FALSE, message = FALSE, echo=FALSE}
# df_mean_coefs_from_sim

df_mean_coefs_from_sim %>%
  kbl(align = "l", caption = "Summary statistics of the estimated SMDs") %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB") %>%
  column_spec(column = 1, border_right = T)

@@

#aggregate_results_overall
aggregate_results_overall %>%
  kbl(caption = "Meta-regression results") %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB") %>%
  column_spec(column = 1, border_right = T)

@@

# We need to read in the appropriate dataset which describes the active and control conditions used in
# psychotherapy trials.

@@

@@

control_summary %>%
  kbl(align = "l", caption = "Comparing the intensity of the intervention between active and control arms of psychotherapy studies", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>% 
  column_spec(1, border_right = T) %>% 
  pack_rows("Number of sessions", 1, 2) %>% 
  pack_rows("Frequency (weeks)", 3, 4) %>% 
  pack_rows("Session length (mins)", 5, 6) %>%
  pack_rows("Total intervention hours", 7, 8) %>%
  pack_rows("Total intervention period (weeks)", 9, 10)

@@

@@

control_summary_t_tests %>%
  kbl(align = "l", caption = "Results for t-tests comparing intervention intensity between active and control arms of psychotherapy trials") %>% 
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>% 
  row_spec(row = 0, background = "#EBEBEB") %>% 
  column_spec(column = 1, border_right = T)

Overall, the active arms of psychotherapy studies were considerably more intensive than the control arms they were compared against (see Table 6 for summary statistics). Active intervention arms featured significantly more sessions than control arms (*t* (`r control_summary_t_tests$df[control_summary_t_tests$Outcome == "Number of sessions"]`) = `r control_summary_t_tests$"t statistic"[control_summary_t_tests$Outcome == "Number of sessions"]`, *p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Number of sessions"]`). These sessions were also significantly longer (*t* (`r control_summary_t_tests$df[control_summary_t_tests$Outcome == "Session length (mins)"]`) = `r control_summary_t_tests$"t statistic"[control_summary_t_tests$Outcome == "Session length (mins)"]`, *p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Session length (mins)"]`) and more frequent (*t* (`r control_summary_t_tests$df[control_summary_t_tests$Outcome == "Frequency (weeks)"]`) = `r control_summary_t_tests$"t statistic"[control_summary_t_tests$Outcome == "Frequency (weeks)"]`, *p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Frequency (weeks)"]`). The total hours involved in an intervention were higher in active vs control arms (*t* (`r control_summary_t_tests$df[control_summary_t_tests$Outcome == "Total intervention hours"]`) = `r control_summary_t_tests$"t statistic"[control_summary_t_tests$Outcome == "Total intervention hours"]`, *p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Total intervention hours"]`), though the total period of the intervention was similar (*t* (`r control_summary_t_tests$df[control_summary_t_tests$Outcome == "Total intervention period (weeks)"]`) = `r control_summary_t_tests$"t statistic"[control_summary_t_tests$Outcome == "Total intervention period (weeks)"]`, *p* = `r control_summary_t_tests$"p-value"[control_summary_t_tests$Outcome == "Total intervention period (weeks)"]`).

@@

```{r, warning=FALSE, message = FALSE, echo=FALSE}
df_severity_sens %>%
  mutate_all(~if_else(is.na(.), "", as.character(.))) %>% 
  kbl(align = "l", caption = "Table S1: Sensitivity analyses for comparison of baseline severity levels between psychotherapy and medication trials", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>% 
  row_spec(row = 1, bold = T) %>% 
  row_spec(c(4, 7), bold = T, extra_css = "border-top: 1px solid black;") %>%
  column_spec(1, border_right = T) %>% 
  add_indent(c(2, 3, 5, 6, 8, 9)) %>% 
  footnote(number = c("Statistic derived from the test for subgroup differences in the relevant meta-analysis"))
```

@@

```{r, warning=FALSE, message = FALSE, echo=FALSE}
df_gender_sens %>%
  mutate_all(~if_else(is.na(.), "", as.character(.))) %>% 
  kbl(align = "l", caption = "Table S2: Sensitivity analyses for comparison of gender composition between psychotherapy and medication trials", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>% 
  row_spec(row = 1, bold = T) %>% 
  row_spec(c(4, 7), bold = T, extra_css = "border-top: 1px solid black;") %>%
  column_spec(1, border_right = T) %>% 
  add_indent(c(2, 3, 5, 6, 8, 9)) %>% 
  footnote(number = c("Statistic derived from the test for subgroup differences in the relevant meta-analysis"))
```

@@

```{r, warning=FALSE, message = FALSE, echo=FALSE}
df_age_sens %>%
  mutate_all(~if_else(is.na(.), "", as.character(.))) %>% 
  kbl(align = "l", caption = "Table S3: Sensitivity analyses for comparison of age between psychotherapy and medication trials", escape = F) %>%
  kable_material(lightable_options = c("hover"), full_width = FALSE) %>%
  row_spec(row = 0, background = "#EBEBEB", extra_css = "border-bottom: 1px black;" ) %>% 
  row_spec(row = 1, bold = T) %>% 
  row_spec(c(4, 7), bold = T, extra_css = "border-top: 1px solid black;") %>%
  column_spec(1, border_right = T) %>% 
  add_indent(c(2, 3, 5, 6, 8, 9)) %>% 
  footnote(number = c("Statistic derived from the test for subgroup differences in the relevant meta-analysis"))
```
